### 2021-04-30 submission/manuscript.Rmd
Edit results section

## Abstract
(Modify depending on target journal, currently abstract submitted to World Microbe Forum) Antibiotics are a major risk factor for *Clostridioides difficile* infections (CDIs) because of their impact on the intestinal microbiome. However, non-antibiotic medications such as the ubiquitous osmotic laxative polyethylene glycol (PEG) 3350, also alter the microbiota, but whether PEG impacts CDI susceptibility and clearance is unclear. To examine how PEG impacts susceptibility, we treated C57Bl/6 mice with 5-day and 1-day doses of 15% PEG in the drinking water and then challenged the mice with *C. difficile* 630 spores. We used clindamycin-treated mice as a control because they consistently clear *C. difficile* within 10 days post-infection (dpi). To examine how PEG treatment impacts clearance, we administered PEG for 1 day to clindamycin-treated, *C. difficile*-challenged mice either immediately following challenge or 3 dpi. We collected longitudinal stool samples to examine *C. difficile* levels in the stool via anaerobic culture and profiled the microbiota by 16S rRNA sequencing. PEG treatment alone was sufficient to render mice susceptible to CDI and 5-day PEG-treated mice remain colonized for up to 30 dpi. Additionally, 5-day PEG treated mice remained susceptible to CDI 10-days post treatment. In contrast, 1-day PEG treated mice were transiently colonized, clearing *C. difficile* within 7 dpi. Although 5-day PEG-treated mice exhibited prolonged *C. difficile* colonization, we saw no difference in histological inflammation between PEG- and clindamycin-treated mice. Additionally, administering PEG to mice after *C. difficile* challenge prolonged colonization up to 30 dpi in mice that received PEG immediately after challenge and 15 dpi in mice that received PEG 3 dpi. When we examined microbiota composition across our different treatment groups, we found decreased richness in the PEG-treated mice that exhibited prolonged *C. difficile* colonization. Importantly, there were increased Bacteroides and Enterobacteriaceae and decreased Lachnospiraceae and Oscillibacter in most of the PEG-treated mice with prolonged *C. difficile* colonization. Our findings suggest the osmotic laxative PEG 3350 alters the mouse microbiota and disrupts colonization resistance to *C. difficile*, as well as clearance in mice with a CDI. Considering that most hospitals recommend not performing *C. difficile* testing on patients taking laxatives and laxatives are used when administering fecal microbiota transplants via colonoscopy to patients with recurrent CDIs, further studies are needed to evaluate if laxatives impact human microbiota colonization resistance.

@@

Many of the non-antibiotic medications associated with CDIs are known to modulate gastrointestinal motility leading to either increased or decreased colonic transit time, which in turn also strongly impacts microbiota composition and function [@Vandeputte2015; @VujkovicCvijin2020]. Stool consistency often serves as an approximation of intestinal motility. Our group has shown that when *C. difficile* negative controls are separated into two groups based on stool consistency, there are shared microbiota features such as lower alpha diversity in samples from CDI patients and control patients with diarrhea compared to control samples that were *C. difficile* negative with non-diarrheal consistency [@Schubert2014]. These results led to a hypothesis that bacterial communities from patients experiencing diarrhea are susceptible to developing CDIs.

Osmotic laxatives can lead to diarrhea depending on the administered dose and temporarily disrupt the human intestinal microbiota [@Nagata2019]. The ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 is found in Miralax, Nulytely, and Golytely and is also commonly used as bowel preparation for colonoscopies. Interestingly, previous studies have shown that treating mice with PEG alone altered microbiota composition, reduced acetate and butyrate production, altered the mucus barrier, and rendered the mice susceptible to *C. difficile* infection [@Kashyap2013; @Ferreyra2014; @Tropini2018; @VanInsberghe2020]. The mucus barrier is thought to mediate protection from *C. difficile* infections by protecting intestinal epithelial cells from the toxins produced by *C. difficile* [@Olson2013; @Diebel2014]. However, whether laxative administration results in more severe CDIs in mice and how long mice remain colonized with *C. difficile* after challenge is unclear.

@@

In the past, our group has used C57BL6 mice to characterize how antibiotics including clindamycin disrupt the microbiota and influence *C. difficile* susceptibility and clearance [@Schubert2015; @Jenior2017; @Tomkovich2020]. Although, two groups have now shown PEG treatment alone renders mice susceptible to *C. difficile* [@Ferreyra2014; @VanInsberghe2020], these studies have raised additional questions regarding the dynamics and severity of infection as well as the role of laxative treatment in *C. difficile* clearance that should be addressed to better inform how we think about laxatives in the context of CDIs. Here, we used our C57BL/6 clindamycin model as a control group to characterize how long PEG-treated mice remain susceptible, whether PEG treatment results in more severe CDI and sustained *C. difficile* colonization, and if PEG treatment post-CDI can promote *C. difficile* clearance.

@@

**5-day laxative treatment leads to prolonged *C. difficile* colonization in mice.** We compared mice treated with the osmotic laxative PEG 3350 to our standard 10 mg/kg clindamycin treatment, which temporarily renders the mice susceptible to *C. difficile*, with mice typically clearing *C. difficile* within 10 days post-infection [@Tomkovich2019; @Tomkovich2020]. All PEG-treated mice were administered a 15% PEG solution in the drinking water for 5-days, one group was also treated with clindamycin, and one group was allowed to recover for 10 days prior to challenge (Fig. 1A). PEG treatment resulted in weight loss in all 3 groups of PEG-treated mice, with the greatest change in weight observed on the fifth day of PEG treatment and the mice recovered most of the weight five days after treatment (Fig. 1B). After either PEG, clindamycin, or PEG and clindamycin treatment all mice were challenged with 10^5^ *C. difficile* 630 spores (Fig. 1A). All treatments rendered mice susceptible to *C. difficile* colonization. However, PEG-treated mice remained colonized with *C. difficile* at a high level through thirty days post-infection (Fig. 1C). In contrast, the clindamycin-treated mice cleared *C. difficile* within ten days post-infection. Surprisingly, PEG-treated mice were still susceptible to *C. difficile* infection after 10-days of recovery from treatment, although *C. difficile* was not detectable in most of the group in the initial five days post-infection (Fig. 1C). From 8 days post-infection onward, the median *C. difficile* CFU stabilized for the 10-day recovery group of PEG-treated mice and remained high through 30 days post-infection (Fig. 1C). Thus, osmotic laxative treatment alone was sufficient to render mice susceptible to prolonged *C. difficile* colonization and PEG-treated mice remained susceptible up to ten days post-treatment.

@@

```
**5-day laxative treatment differentially disrupts the fecal microbiota compared to clindamycin treatment.** Since osmotic laxatives and clindamycin have previously been shown to disrupt the murine microbiota [@Kashyap2013; @Ferreyra2014; @Tropini2018; @VanInsberghe2020], we hypothesized the different *C. difficile* colonization dynamics between mice treated with the osmotic laxative or clindamycin were due to the two drugs having differential effects on the microbiota. We profiled the stool microbiota over time by sequencing the V4 region of the 16S rRNA gene to compare changes across treatment groups. We found time and treatment group explained half of the observed variation between fecal communities with most of the remaining variation explained by interactions between treatment group and other experimental variables including time, cage effects, and sequencing preparation plate (PERMANOVA combined R^2^ = `r f_d_top_R2_sum`, *P* `r format.pval(f_d_top_p, eps = 0.001)`, Fig. 2A, Data Set S1, Sheet X). Cage effects refer to the well-documented phenomenon that mice housed in the same cages have similar microbial communities due to copraphagy [@Nguyen2015]. We tried to minimize the impact of cage effects on our experiment by breaking up cagemates when assigning mice to treatment groups and primarily housing only two mice per cage. Importantly, although we conducted a total of `r five_day_total_exp` separate experiments, the experiment number and it's interaction with treatment group was not one of the variables that significantly explained the observed variation in fecal communities (Data Set S1, Sheet X). Interestingly, none of the treatment groups recovered to their baseline community structure either 10 or 30 days post-infection suggesting other community features besides recovery to baseline were responsible for the prolonged *C. difficile* colonization in PEG-treated mice (Fig. 2B).

Next, we examined  alpha diversity by looking at Shannon diversity over time. Although both clindamycin and PEG treatments decreased diversity, Shannon diversity was lower in the groups of mice that received PEG treatment compared to those that received clindamycin through thirty days post-infection (Fig. 2C). We next examined the bacterial genera that shifted after PEG treatment by comparing the baseline samples of mice treated with only PEG to samples from the same mice one day post-treatment. We found `r PEG_genera` genera that were altered by PEG treatment (Data Set S1, Sheet X). The majority of the bacterial relative abundances decreased after PEG, but *Enterobacteriaceae* and *Bacteroides* increased and the increase in *Bacteroides* was unique to PEG treatment, as *Bacteroides* actually decreased in clindamycin treated mice (Fig. 2D). Finally, we examined the bacteria that differed across treatment groups over multiple time points. We found `r f_d_group_multi_day` genera were different over multiple time points out of the `r f_d_group_total_genera` genera that were different between treatment groups (Fig. 2E, Data Set S1, Sheet X). Thus, PEG had a significant impact on the fecal microbiota that was maintained over time and was distinct from clindamycin treatment.

Interestingly, *C. difficile* was not immediately detectable in the stools of the PEG-treated mice that were allowed to recover for 10 days prior to  challenge. We decided to examine the bacteria that changed during the post-infection period when the group median *C. difficile* shifted from undetectable at 1 day post-infection to detectable in the stool samples with the median stabilizing around 8 days post-infection (Fig. S1A). Interestingly, we found *Erysipelotrichaceae*, *Enterobacteriaceae*, and *Akkermansia* were changing during the time period when *C. difficile* was becoming detectable in the stools (Fig. S1B), although none of the bacteria were significant after multiple hypothesis correction (Data Set S1, Sheet X). While we did not identify a clear signal to explain the delayed appearance of *C. difficile* in the 5-day PEG mice that were allowed to recover for 10 days prior to challenge, the delay is striking and could reflect changes in microbial activity or metabolites that were not examined in this study.

@@

```
**5-day laxative treatment does not promote more severe CDIs despite altering the mucosal microbiota.** Given the findings from a previous study that demonstrated PEG treatment disrupts the mucus layer and alters the immune response in mice [@Tropini2018], we decided to examine the impact of PEG treatment on the mucosal microbiota and CDI severity. To evaluate the mucosal microbiota, we sequenced snips of tissue collected from the cecum, proximal colon, and distal colon. Similar to what was observed with the stool samples, alpha diversity was lower in the PEG-treated mice compared to clindamycin treatment (Fig. 3A). Although alpha diversity continued to increase over time based on the communities from PEG-treated mice collected at 20 and 30 days post-infection (Fig. 3A, Data Set S1, Sheet X). Group, time point, and their interactions with other variables (cage, experiment number, and sample type) explained the majority of the variation observed in mucosal communities (PERMANOVA combined R^2^ = `r f_d_top_R2_sum_t`, *P* `r format.pval(f_d_top_p_t, eps = 0.05)`, Fig. 3B, Data Set S1, Sheet X). We saw the greatest difference in the mucosal microbiota between treatment groups at 6 days post-infection with `r shared_tissue_genera` genera that were significantly different in all three of the tissue types we collected (cecum, proximal colon, and distal colon; Fig. S3A, Data Set S1, Sheet X). However, at 30 days post-infection only *Bacteroides*, *Clostridiales*, *Firmicutes*, and *Ruminococcaceae* were different between treatment groups and only in the cecum tissues (Fig. 3C, Fig. 2E, Data Set S1, Sheet X). Interestingly, one of the bacteria that was different between treatment groups at 6 days post-infection was *Peptostreptococcaceae* (the genera that contains a sequence that matches *C. difficile*). *Peptostreptococcaceae* was primarily only present in the 5-day PEG treatment group of mice and decreased in the proximal and distal colon tissues over time (Fig. S2B). Thus, PEG treatment had a significant impact on the mucosal microbiota and we detected *C. difficile* sequences in the cecum, proximal colon, and distal colon tissue communities.

Next, we examined CDI severity by evaluating cecum and colon histopathology [@Reeves2011] and found there was no difference in cecum and colon scores between clindamycin and PEG-treated mice that were challenged with *C. difficile* at 4 days post-infection (Fig. 3D), the time point typically examined in *C. difficile* 630 challenged mice [@Dieterle2020]. We also looked at 6 days post-infection because that was when we started to see a large difference in *C. difficile* colonization levels between PEG- and clindamycin-treated mice (Fig. 1C). Although, there was a slight difference in the colon between PEG and clindamycin-treated mice, there was no difference in the cecum and the overall score is still relatively low given that the max possible summary score is twelve (Fig. 3E). Therefore, although PEG treatment had a disruptive effect on the mucosal microbiota, the impact of *C. difficile* 630 infection on the cecum and colon was similar between PEG and clindamycin treated mice.

***C. difficile* challenge does not have a synergistic disruptive effect on the microbiota of PEG-treated mice** Because *C. difficile* itself can have an impact on the microbiota [@Jenior2018], we also sequenced the tissue and stools of mock-challenged clindamycin and 5-day PEG treated mice. Examining the stools of the mock-challenged mice revealed similar bacterial disruptions as the *C. difficile* challenged mice (Fig. S2A-C). Similarly, there was no difference between the tissues of mock and *C. difficile* challenged mice (Fig. S2D-F). Thus, most of the microbiota alterations we observed in the PEG-treated mice were a result of the laxative and not an interaction between the laxative and *C. difficile*.
``` {r 1-day PEG stool microbiota}

#Genera that were different between treatment days
#Genera that were changed by treatment
pt_d1 <- read_tsv(here("data/process/1_Day_PEG_genus_stats_BtoD1.tsv")) %>%
  filter(p.value.adj < 0.05)
#Genera that were still different at 7 days post-infection  
pt_d7 <- read_tsv(here("data/process/1_Day_PEG_genus_stats_BtoD7.tsv")) %>%
  filter(p.value.adj < 0.05)
pt_d7_genera <- pt_d7 %>% pull(genus)

@@

```
**1-day laxative treatment results in transient *C. difficile* colonization and minor microbiota disruption** Next, we decided to examine how a shorter osmotic laxative perturbation would impact the microbiome and susceptibility to *C. difficile*. We administered either a 1-day PEG treatment, a 1-day PEG treatment with a 1-day recovery period, or clindamycin to mice before challenging them with *C. difficile* (Fig. 3A). In contrast to the 5-day PEG treated mice, the 1-day PEG treated mice were only transiently colonized and cleared *C. difficile* by 7 days post-infection (Fig. 3B). The stool communities of PEG-treated mice were also only transiently disrupted, with Shannon diversity recovering close to baseline by 7 days post-infection (Fig. 3C-D). We found `r recovered_genera` bacteria were impacted by treatment, but recovered close to baseline levels by 7 days post-infection including *Enterobacteriaceae*, *Clostridiales*, *Porpyromonadaceae*, and *Ruminococcaceae* (Fig. 3E, Data Set S1, Sheet X). These findings suggest the duration of treatment matters when considering the impact of the laxatives, with shorter treatments resulting in a transient loss of *C. difficile* colonization resistance.

@@

```
**Post-CDI laxative treatment disrupts clearance in clindamycin-treated mice regardless of whether an FMT is also administered** Since 1-day PEG treatment resulted in a more mild microbiota perturbation, we decided to use the 1-day treatment to examine the hypothesis that PEG helps to flush *C. difficile* spores from the intestine. We were also interested in exploring whether PEG might help with engraftment in the context of FMTs. To examine the impact of PEG treatment on *C. difficile* clearance and FMT treatment, we treated 4 groups of mice with clindamycin and then challenged all mice with *C. difficile* before administering the following treatments: no additional treatment, 1-day PEG immediately after challenge, and 1-day PEG treatment 3 days after challenge followed by either administration of an FMT or PBS solution by oral gavage (Fig. 5A). Contrary to our hypothesis, all groups of mice that received PEG exhibited prolonged *C. difficile* colonization (Fig. 5B). Next we examined how post-CDI PEG treatment impacted the microbiota. Alpha diversity was lower in the PEG-treated mice with the exception of the PEG-treated mice that were administered an FMT (Fig 5C-D). The FMT appeared to partially restore Shannon diversity but not richness (Fig. 5C-D). Similarly, we saw some overlap between the communities of mice that received FMT and the mice treated with only clindamycin after 5 days post-infection (Fig. 6A). The increase in Shannon diversity suggests that the FMT did have an impact on the microbiota, despite seeing prolonged *C. difficile* colonization in the FMT treated mice. However, only *Bacteroidales* and *Porphyromonadaceae* consistently differed between the mice received either an FMT or PBS gavage (Fig. 6B), suggesting the FMT only restored a couple of genera. Overall, we found there were `r post_group_multi_day` bacteria that differed over multiple time points between treatment groups including increased *Akkermansia* and decreased *Ruminococcaceae*, *Clostridiales*, *Lachnospiraceae*, and *Oscillibacter* in mice that received PEG after *C. difficile* challenge (Fig. 6C). In sum, administering PEG actually prolonged *C. difficile* colonization, including in mice that received an FMT, which only restored a couple of bacterial genera.

>>>>

## Abstract
(Modify depending on target journal, currently abstract submitted to World Microbe Forum) Antibiotics are a major risk factor for *Clostridioides difficile* infections (CDIs) because of their impact on the intestinal microbiome. However, non-antibiotic medications such as the ubiquitous osmotic laxative polyethylene glycol (PEG) 3350, also alter the microbiota, but whether PEG impacts CDI susceptibility and clearance is unclear. To examine how PEG impacts susceptibility, we treated C57Bl/6 mice with 5-day and 1-day doses of 15% PEG in the drinking water and then challenged the mice with *C. difficile* 630 spores. We used clindamycin-treated mice as a control because they consistently clear *C. difficile* within 10 days post-challenge (dpc). To examine how PEG treatment impacts clearance, we administered PEG for 1 day to clindamycin-treated, *C. difficile*-challenged mice either immediately following challenge or 3 dpc. We collected longitudinal stool samples to examine *C. difficile* levels in the stool via anaerobic culture and profiled the microbiota by 16S rRNA sequencing. PEG treatment alone was sufficient to render mice susceptible to CDI and 5-day PEG-treated mice remain colonized for up to 30 dpc. Additionally, 5-day PEG treated mice remained susceptible to CDI 10-days post treatment. In contrast, 1-day PEG treated mice were transiently colonized, clearing *C. difficile* within 7 dpc. Although 5-day PEG-treated mice exhibited prolonged *C. difficile* colonization, we saw no difference in histological inflammation between PEG- and clindamycin-treated mice. Additionally, administering PEG to mice after *C. difficile* challenge prolonged colonization up to 30 dpc in mice that received PEG immediately after challenge and 15 dpc in mice that received PEG 3 dpc. When we examined microbiota composition across our different treatment groups, we found decreased richness in the PEG-treated mice that exhibited prolonged *C. difficile* colonization. Importantly, there were increased Bacteroides and Enterobacteriaceae and decreased Lachnospiraceae and Oscillibacter in most of the PEG-treated mice with prolonged *C. difficile* colonization. Our findings suggest the osmotic laxative PEG 3350 alters the mouse microbiota and disrupts colonization resistance to *C. difficile*, as well as clearance in mice with a CDI. Considering that most hospitals recommend not performing *C. difficile* testing on patients taking laxatives and laxatives are used when administering fecal microbiota transplants via colonoscopy to patients with recurrent CDIs, further studies are needed to evaluate if laxatives impact human microbiota colonization resistance.

@@

Many of the non-antibiotic medications associated with CDIs are known to modulate gastrointestinal motility leading to either increased or decreased colonic transit time, which in turn also strongly impacts microbiota composition and function [@Vandeputte2015; @VujkovicCvijin2020]. Stool consistency often serves as an approximation of intestinal motility [@Vandeputte2015]. Our group has shown that when *C. difficile* negative controls are separated into two groups based on stool consistency, there are shared microbiota features such as lower alpha diversity in samples from CDI patients and control patients with diarrhea compared to control samples that were *C. difficile* negative with non-diarrheal consistency [@Schubert2014]. These results led to a hypothesis that bacterial communities from patients experiencing diarrhea are susceptible to developing CDIs.

Osmotic laxatives can lead to diarrhea depending on the administered dose and temporarily disrupt the human intestinal microbiota [@Nagata2019]. The ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 is found in Miralax, Nulytely, and Golytely and is also commonly used as bowel preparation for colonoscopies. Interestingly, previous studies have shown that treating mice with PEG alone altered microbiota composition, reduced acetate and butyrate production, altered the mucus barrier, and rendered the mice susceptible to *C. difficile* infection [@Kashyap2013; @Ferreyra2014; @Tropini2018; @VanInsberghe2020]. The mucus barrier is thought to mediate protection from *C. difficile* infections by protecting intestinal epithelial cells from the toxins produced by *C. difficile* [@Olson2013; @Diebel2014]. Whether laxative administration results in more severe CDIs in mice and how long mice remain colonized with *C. difficile* after challenge is unclear.

@@

Our group has used C57BL/6 mice to characterize how antibiotics disrupt the microbiota and influence *C. difficile* susceptibility and clearance [@Schubert2015; @Jenior2017; @Tomkovich2020]. Although, two groups have now shown PEG treatment alone renders mice susceptible to *C. difficile* [@Ferreyra2014; @VanInsberghe2020], these studies have raised additional questions regarding the dynamics and severity of infection as well as the role of laxative treatment in *C. difficile* clearance. Addressing these questions will better inform how we think about laxatives in the context of CDIs. Here, we characterized how long PEG-treated mice remain susceptible, whether PEG treatment results in more sustained *C. difficile* colonization and severe CDI and  than mice treated with clindamycin, and if PEG treatment after colonization can promote *C. difficile* clearance.

@@

**5-day laxative treatment led to prolonged *C. difficile* colonization in mice.** Building off of previous work that showed treating mice with the osmotic laxative, PEG 3350, rendered mice susceptible to *C. difficile* colonization [@Ferreyra2014; @VanInsberghe2020], we decided to test how long *C. difficile* colonization is sustained and how long PEG-treated mice remain susceptible to *C. difficile*.  We compared three groups of mice treated with PEG 3350 to one group of mice treated with our standard 10 mg/kg clindamycin treatment, which temporarily renders the mice susceptible to *C. difficile* colonization, with mice typically clearing *C. difficile* within 10 days post-challenge [@Tomkovich2019; @Tomkovich2020]. All three groups of PEG-treated mice were administered a 15% PEG solution in the drinking water for 5-days, 1. received no additional treatment, 2. was also treated with clindamycin, and 3. was allowed to recover for 10 days prior to challenge (Fig. 1A). PEG treatment resulted in weight loss in all 3 groups of PEG-treated mice relative to their baseline weights, with the greatest change in weight observed on the fifth day of PEG treatment. The mice recovered most of the lost weight by five days after treatment (Fig. 1B). After either the PEG, clindamycin, or PEG and clindamycin treatment all mice were challenged with 10^5^ *C. difficile* 630 spores (Fig. 1A). All treatments rendered mice susceptible to *C. difficile* colonization. However, PEG-treated mice remained colonized with *C. difficile* at a high level through thirty days post-challenge (Fig. 1C). In contrast, the clindamycin-treated mice cleared *C. difficile* within ten days post-challenge (Fig. 1C). Therefore, PEG treatment led to sustained colonization in contrast to clindamycin mice that naturally cleared *C. difficile* within ten days post-challenge.

Notably, we also found PEG-treated mice were still susceptible to *C. difficile* colonization after a 10-day recovery period, although *C. difficile* was not detectable in most of the group in the initial five days post-challenge (Fig. 1C, S1A). One mouse was found dead on the 6th day post-challenge, presumable due to *C. difficile* as the bacteria became detecable in that mouse on the 4th day post-challenge (Fig. S1A, mouse 10). From 8 days post-challenge onward, the density of *C. difficile* stabilized in the 10-day recovery group and remained high through 20-30 days post-challenge (Fig. 1C). Thus, osmotic laxative treatment alone was sufficient to render mice susceptible to prolonged *C. difficile* colonization and PEG-treated mice remained susceptible through ten days post-treatment.

@@

```
**5-day laxative treatment differentially disrupted the fecal microbiota compared to clindamycin treatment.** Since osmotic laxatives and clindamycin have previously been shown to disrupt the murine microbiota [@Kashyap2013; @Ferreyra2014; @Tropini2018; @VanInsberghe2020], we hypothesized the different *C. difficile* colonization dynamics between mice treated with the osmotic laxative or clindamycin were due to the two drugs having differential effects on the microbiota. We profiled the stool microbiota over time by sequencing the V4 region of the 16S rRNA gene to compare changes across treatment groups. We found time and treatment group explained half of the observed variation between fecal communities with most of the remaining variation explained by interactions between treatment group and other experimental variables including time, cage effects, and sequencing preparation plate (PERMANOVA combined R^2^ = `r f_d_top_R2_sum`, *P* `r format.pval(f_d_top_p, eps = 0.001)`, Fig. 2A, Data Set S1, Sheet X). Cage effects refer to the well-documented phenomenon that mice housed in the same cages have similar microbial communities due to coprophagy [@Nguyen2015]. We tried to minimize the impact of cage effects on our experiment by breaking up cagemates when assigning mice to treatment groups. Importantly, although we conducted a total of `r five_day_total_exp` separate experiments, the experiment number and its interaction with treatment group was not one of the variables that significantly explained the observed variation in fecal communities (Data Set S1, Sheet X). None of the treatment groups recovered to their baseline community structure either 10 or 30 days post-challenge suggesting other community features besides recovery to baseline were responsible for the prolonged *C. difficile* colonization in PEG-treated mice (Fig. 2B).

Since none of the communities completely recovered in the follow-up period after treatments, we next profiled community diversity and composition. We examined the alpha diversity dynamics by calculating the communities' Shannon diversity. Although both clindamycin and PEG treatments decreased diversity, Shannon diversity was lower in the groups of mice that received PEG treatment compared to those that received clindamycin through thirty days post-challenge (Fig. 2C). We next examined the bacterial genera that shifted after PEG treatment by comparing the baseline samples of mice treated with only PEG to samples from the same mice one day post-PEG-treatment. We found `r PEG_genera` genera that were altered by PEG treatment (Data Set S1, Sheet X). The majority of the bacterial relative abundances decreased after the PEG treatment, but the relative abundance among members of the *Enterobacteriaceae* and *Bacteroides* increased. The increase in *Bacteroides* relative abundance was unique to PEG treated mice, as the *Bacteroides* relative abundance actually decreased in clindamycin treated mice (Fig. 2D). Finally, we identified the genera whose relative abundance differed across treatment groups over multiple time points. Of the `r f_d_group_total_genera` genera that were different between treatment groups, `r f_d_group_multi_day` genera were different over multiple time points  (Fig. 2E, Data Set S1, Sheet X). Thus, PEG had a significant impact on the fecal microbiota that was maintained over time and was distinct from clindamycin treatment.

Interestingly, *C. difficile* was not immediately detectable in the stools of the PEG-treated mice that were allowed to recover for 10 days prior to  challenge. We decided to examine if there were genera that changed during the post-challenge period when the group median *C. difficile* shifted from undetectable at 1 day post-challenge to detectable in the stool samples with the density stabilizing around 8 days post-challenge (Fig. S1A). We found no bacteria were significantly different over the two time periods after multiple hypothesis correction (Data Set S1, Sheet X). However, there was also wide variation between individual mice regarding when *C. difficile* became detectable (Fig. S1A) as well as the relative abundances of bacterial genera in the communities (Fig. S1B). For example, two mice had a high relative abundance of *Enterobacteriaceae* throughout the post-challenge period and this corresponded to mouse 10, which died on the sixth day post-challenge and mouse 11, where *C. difficile* was present at a high density from the 4th day post-challenge onward (Fig. S1B). While we did not identify a clear signal to explain the delayed appearance of *C. difficile* in the 5-day PEG mice that were allowed to recover for 10 days prior to challenge, the delay is striking and could reflect changes in microbial activity or metabolites that were not examined in this study.

@@

tissue_genera_max_p <- read_tsv(here("data/process/5_days_PEG_genus_group_cecum.tsv")) %>% 
  add_row(read_tsv(here("data/process/5_days_PEG_genus_group_proximal_colon.tsv"))) %>% 
  add_row(read_tsv(here("data/process/5_days_PEG_genus_group_distal_colon.tsv"))) %>% 
  filter(p.value.adj < 0.05) %>% 
  #Limit to genus found in all 3 types of tissues
  filter(genus %in% intersect_all(kw_genus_cecum, kw_genus_proximal_colon, kw_genus_distal_colon)) %>%
  select(p.value.adj) %>% max()

```
**5-day laxative treatment did not promote more severe CDIs despite altering the mucosal microbiota.** Given the findings from a previous study that demonstrated PEG treatment disrupts the mucus layer and alters the immune response in mice [@Tropini2018], we decided to examine the impact of PEG treatment on the mucosal microbiota and CDI severity. To evaluate the mucosal microbiota, we sequenced snips of tissue collected from the cecum, proximal colon, and distal colon. Similar to what was observed with the stool samples, alpha diversity was lower in the PEG-treated mice compared to clindamycin treated mice (Fig. 3A). Alpha diversity continued to increase over time with the PEG-treated mice collected at 20 and 30 days post-challenge (Fig. 3A, Data Set S1, Sheet X). Group, time point, and their interactions with other variables (cage, experiment number, and sample type) explained the majority of the variation observed in mucosal communities (PERMANOVA combined R^2^ = `r f_d_top_R2_sum_t`, *P* `r format.pval(f_d_top_p_t, eps = 0.05)`, Fig. 3B, Data Set S1, Sheet X). We saw the greatest difference in the relative abundance of the mucosal microbiota between treatment groups (clindamycin, 5-day PEG, and 5-day PEG plus clindamycin) at 6 days post-challenge with `r shared_tissue_genera` genera that were significantly different (*P* `r format.pval(tissue_genera_max_p, eps = 0.05)) in all three of the tissue types we collected (cecum, proximal colon, and distal colon; Fig. S2A, Data Set S1, Sheet X). Interestingly, *Peptostreptococcaceae* (the family with a sequence that matches *C. difficile*) was one of the genera that had a significant difference in relative abundance between treatment groups at 6 days post-challenge. This population was primarily only present in the 5-day PEG treatment group of mice and decreased in the proximal and distal colon tissues over time (Fig. S2B). By 30 days post-challenge, only the relative abundances of *Bacteroides*, *Clostridiales*, *Firmicutes*, and *Ruminococcaceae* were different between treatment groups and only in the cecum tissues (Fig. 3C, Fig. 2E, Data Set S1, Sheet X). Thus, PEG treatment had a significant impact on the mucosal microbiota and we detected *C. difficile* sequences in the cecum, proximal colon, and distal colon tissue communities.

Next, we examined the severity of *C. difficile* challenge by evaluating cecum and colon H&E stained histopathology [@Reeves2011] and found there was no difference in cecum and colon scores between clindamycin and PEG-treated mice that were challenged with *C. difficile* at 4 days post-challenge (Fig. 3D), the time point typically examined in *C. difficile* 630 challenged mice [@Dieterle2020]. We also looked at 6 days post-challenge because that was when there was a large difference in *C. difficile* density between PEG- and clindamycin-treated mice (Fig. 1C). Although, there was a slight difference in the colon between PEG and clindamycin-treated mice, there was not a signifant difference in the cecum and the overall score was relatively low (1.5-2.5 out of 12, Fig. 3E). Therefore, although PEG treatment had a disruptive effect on the mucosal microbiota, the impact of *C. difficile* 630 challenge on the cecum and colon was similar between PEG and clindamycin treated mice.

***C. difficile* challenge did not have a synergistic disruptive effect on the microbiota of PEG-treated mice.** Because *C. difficile* itself can have an impact on the microbiota [@Jenior2018], we also sequenced the tissue and stools of mock-challenged clindamycin and 5-day PEG treated mice. Examining the stools of the mock-challenged mice revealed similar bacterial disruptions as the *C. difficile* challenged mice (Fig. S3A-C). Similarly, there was no difference between the tissues of mock and *C. difficile* challenged mice (Fig. S3D-F). Thus, most of the microbiota alterations we observed in the PEG-treated mice were a result of the laxative and not an interaction between the laxative and *C. difficile*.
``` {r 1-day PEG stool microbiota}

#Genera that were different between treatment days
#Genera that were changed by treatment
pt_d1 <- read_tsv(here("data/process/1_Day_PEG_genus_stats_BtoD1.tsv")) %>%
  filter(p.value.adj < 0.05)
#Genera that were still different at 7 days post-challenge  
pt_d7 <- read_tsv(here("data/process/1_Day_PEG_genus_stats_BtoD7.tsv")) %>%
  filter(p.value.adj < 0.05)
pt_d7_genera <- pt_d7 %>% pull(genus)

@@

```
**1-day laxative treatment resulted in transient *C. difficile* colonization and minor microbiota disruption.** Next, we examined how a shorter osmotic laxative perturbation would impact the microbiome and susceptibility to *C. difficile*. We administered either a 1-day PEG treatment, a 1-day PEG treatment with a 1-day recovery period, or clindamycin to mice before challenging them with *C. difficile* (Fig. 3A). In contrast to the 5-day PEG treated mice, the 1-day PEG treated mice were only transiently colonized and cleared *C. difficile* by 7 days post-challenge (Fig. 3B). The stool communities of PEG-treated mice were also only transiently disrupted, with Shannon diversity recovering by 7 days post-challenge (Fig. 3C-D). We found the relative abundances of `r recovered_genera` genera were impacted by treatment, but recovered close to baseline levels by 7 days post-challenge including *Enterobacteriaceae*, *Clostridiales*, *Porpyromonadaceae*, and *Ruminococcaceae* (Fig. 3E, Data Set S1, Sheet X). These findings suggest the duration of the PEG treatment was relevant, with shorter treatments resulting in a transient loss of *C. difficile* colonization resistance.

@@

```
**Post-CDI laxative treatment disrupted clearance in clindamycin-treated mice regardless of whether an FMT was also administered.** Since a 1-day PEG treatment resulted in a more mild microbiota perturbation, we decided to use the 1-day treatment to examine the hypothesis that PEG helps to flush *C. difficile* spores from the intestine. To examine the impact of PEG treatment on *C. difficile* clearance, we treated 4 groups of mice with clindamycin and then challenged all mice with *C. difficile* before administering the following treatments: no additional treatment, 1-day PEG immediately after challenge, and 1-day PEG treatment 3 days after challenge followed by either administration of an FMT or PBS solution by oral gavage (Fig. 5A). Contrary to our hypothesis, all groups of mice that received PEG exhibited prolonged *C. difficile* colonization (Fig. 5B). Next we examined how post-CDI PEG treatment impacted microbiota diversity. Alpha diversity was lower in the PEG-treated mice with the exception of the PEG-treated mice that were administered an FMT (Fig 5C-D).

We were also interested in exploring whether PEG might help with engraftment in the context of FMTs. The FMT appeared to partially restore Shannon diversity but not richness (Fig. 5C-D). Similarly, we saw some overlap between the communities of mice that received FMT and the mice treated with only clindamycin after 5 days post-challenge (Fig. 6A). The increase in Shannon diversity suggests that the FMT did have an impact on the microbiota, despite seeing prolonged *C. difficile* colonization in the FMT treated mice. However, only the relative abundances of *Bacteroidales* and *Porphyromonadaceae* consistently differed between the mice received either an FMT or PBS gavage (Fig. 6B), suggesting the FMT only restored a couple of genera. Overall, we found the relative abundances of `r post_group_multi_day` genera were different between treatment groups over multiple timepoints. For example, the relative abundance of *Akkermansia* was increased and the relative abundances of *Ruminococcaceae*, *Clostridiales*, *Lachnospiraceae*, and *Oscillibacter* were decreased in mice that received PEG after *C. difficile* challenge relative to clindamycin treated mice (Fig. 6C). In sum, administering PEG actually prolonged *C. difficile* colonization, including in mice that received an FMT, which only restored 2 bacterial genera.
